### 2022-05-27 RFCs/HDF5/VOL/connector_author_guide/writing_new.tex

\subsection{Sources Of Information}

@@

\begin{tabularx}{\linewidth}{ l X }
  \texttt{H5VLpublic.h} & Public VOL header.\\
  \texttt{H5VLconnector.h} & Main connector author header. Contains definitions for the main VOL struct and callbacks, enum values, etc.\\
  \texttt{H5VLconnector\_passthru.h} & Helper routines for passthrough connector authors.\\
    \texttt{H5VLnative.h} & Native VOL connector header. May be useful if your connector will attempt to implement native HDF5 API calls that are handled via the \textit{optional} callbacks.\\
  \texttt{H5PLextern.h} & Needed if your connector will be built as a plugin.\\
\end{tabularx}

Many VOL connectors can be found on The HDF Group's Bitbucket server, located at
\url{https://bitbucket.hdfgroup.org/projects/HDF5VOL}. Not all of these VOL
connectors are supported by THe HDF Group and the level of completeness varies,
but the connectors found there can serve as examples of working implementations.

@@

@@

For the HDF5 library to be able to load an external plugin dynamically, the plugin developer has to define two public routines with the following name and signature:
\begin{lstlisting}
H5PL_type_t H5PLget_plugin_type(void)
const void *H5PLget_plugin_info(void)
\end{lstlisting}
{\tt H5PLget\_plugin\_type} should return the library type which should always be {\tt H5PL\_TYPE\_VOL}. {\tt H5PLget\_plugin\_info} should return a pointer to the plugin structure defining the VOL plugin with all the callbacks. For example, consider an external plugin defined as:
\begin{lstlisting}
static const H5VL_class_t H5VL_log_g = {
    1,	/* version */
    502,	/* value */
    "log",	/* name */
    ...
}
\end{lstlisting}
The plugin would implement the two routines as:
\begin{lstlisting}
H5PL_type_t H5PLget_plugin_type(void) {return H5PL_TYPE_VOL;}
const void *H5PLget_plugin_info(void) {return &H5VL_log_g;}
\end{lstlisting}

@@

We have created a template VOL connector that includes both Autotools and CMake
build files. The constructed VOL connector includes no real functionality, but 
can be registered and loaded as a plugin.

@@

\quad \quad \url{https://bitbucket.hdfgroup.org/projects/HDF5VOL/repos/template}

@@

@@

In HDF5 1.12.0, {\tt version} will be 1, indicating version 1 of the {\tt H5VL\_class\_t} struct.

@@

VOL connector values are integers, with a maximum value of 65535. Values from
0 to 255 are reserved for internal use by The HDF Group. The native VOL connector
has a value of 0.

As is the case with HDF5 filters, The HDF Group can register the value you've
chosen for your VOL connector so that others don't use it. Please contact
\href{mailto:help@hdfgroup.org}{help@hdfgroup.org} for help with this. We currently
do not register connector names.

The {\tt cap\_flags} field is currently not fully utilized. In the future, this
will be most likely be used to indicate which callbacks and capabilities are
supported by the VOL connector, but for now this struct field contains few flags, 
a list of which can be found in {\tt H5VLconnector.h}.

@@

You'll need to decide how to perform any initialization and shutdown tasks that
are required by your connector. There are initialize and terminate callbacks
in the {\tt H5VL\_class\_t} struct to handle this. They are invoked when
the connector is registered and unregistered, respectively. If this is
unsuitable, you may have to create custom connector-specific API calls to handle
initialization and termination. It may also be useful to perform operations in
a custom API call used to set the VOL connector in the fapl.

@@

Once you have your data structures, you'll need to create your own
implementations of the callback functions and map them via your
{\tt H5VL\_class\_t} struct. 
\subsection{Testing Your Connector}

@@

\quad \quad \url{https://bitbucket.hdfgroup.org/projects/HDF5VOL/repos/vol-tests}

@@

>>>>

\subsection{The HDF5 1.12.x VOL Interface Is DEPRECATED}

Important changes were made to the VOL interface for HDF5 1.13.0 and, due to
binary compatibility issues, these cannot be merged to HDF5 1.12.x. For this
reason, VOL connector development should be shifted to target 1.13.0 as no
further development of the VOL interface will take place on the 1.12.x branch.
Unlike the other development branches of the library, there is no
{\tt hdf5\_1\_13} branch - all HDF5 1.13.0 development is taking place in
the {\tt develop} branch of the HDF5 repository and 1.13.x branches will split
off from that.

Note also that HDF5 1.13.0 is considered an unstable branch, and the API and
file format are subject to change ("unstable" means "not yet finalized", not "buggy").
The VOL feature is under active development and, although it is nearing its final form,
may change further before the stable HDF5 1.14.0 release targeted for 2022.

\subsection{VOL-Related HDF5 Header Files}

@@

\begin{tabularx}{\linewidth}{ l X }
  \texttt{H5VLpublic.h} & Public VOL header.\\
  \texttt{H5VLconnector.h} & Main header for connector authors. Contains definitions for the main VOL struct and callbacks, enum values, etc.\\
  \texttt{H5VLconnector\_passthru.h} & Helper routines for passthrough connector authors.\\
    \texttt{H5VLnative.h} & Native VOL connector header. May be useful if your connector will attempt to implement native HDF5 API calls that are handled via the \textit{optional} callbacks.\\
  \texttt{H5PLextern.h} & Needed if your connector will be built as a plugin.\\
\end{tabularx}

Many VOL connectors are listed on The HDF Group's VOL plugin registration page,
located at \url{https://portal.hdfgroup.org/display/support/Registered+VOL+Connectors}.
Not all of these VOL connectors are supported by The HDF Group and the level
of completeness varies, but the connectors found there can serve as examples
of working implementations.

@@

For the HDF5 library to be able to load an external plugin dynamically, the plugin developer has to define two public routines with the following name and signature:
\begin{lstlisting}
H5PL_type_t H5PLget_plugin_type(void)
const void *H5PLget_plugin_info(void)
\end{lstlisting}

@@

{\tt H5PLget\_plugin\_type} should return the library type which should always be {\tt H5PL\_TYPE\_VOL}. {\tt H5PLget\_plugin\_info} should return a pointer to the plugin structure defining the VOL plugin with all the callbacks. For example, consider an external plugin defined as:

\begin{lstlisting}
static const H5VL_class_t H5VL_foo_g = {
    2,      /* version */
    12345,  /* value */
    "foo",  /* name */
    ...
}
\end{lstlisting}

The plugin would implement the two routines as:
\begin{lstlisting}
H5PL_type_t H5PLget_plugin_type(void) {return H5PL_TYPE_VOL;}
const void *H5PLget_plugin_info(void) {return &H5VL_foo_g;}
\end{lstlisting}

@@

We have created a template terminal VOL connector that includes both Autotools and CMake
build files. The constructed VOL connector includes no real functionality, but 
can be registered and loaded as a plugin.

@@

\quad \quad \url{https://github.com/HDFGroup/vol-template}

@@

A template passthrough VOL is also available. This will be discussed in the
section on passthrough connectors.

@@

In HDF5 1.13.0, the {\tt version} field will be 2, indicating the connector
targets version 2 of the {\tt H5VL\_class\_t} struct. Version 1 of the struct
was never formally released and only available in the {\tt develop} branch of
the HDF5 git repository. Version 0 is used in the deprecated HDF5 1.12.x branch.

@@

VOL connector values are integers, with a maximum value of 65535. Values from
0 to 255 are reserved for internal use by The HDF Group. The native VOL connector
has a value of 0. Values of 256 to 511 are for connector testing and should
not be found in the wild. Values of 512 to 65535 are for external connectors.

As is the case with HDF5 filters, The HDF Group can assign you an official
VOL connector value. Please contact
\href{mailto:help@hdfgroup.org}{help@hdfgroup.org} for help with this. We currently
do not register connector names, though the name you've chosen will appear on
the registered VOL connectors page.

As noted above, registered VOL connectors will be listed at:

\url{https://portal.hdfgroup.org/display/support/Registered+VOL+Connectors}

A new {\tt conn\_version} field has been added to the class struct for 1.13.
This field is currently not used by the library so its use is determined by
the connector author. Best practices for this field will be determined in
the near future and this part of the guide will be updated.

The {\tt cap\_flags} field is used to determine the capabilities of the VOL
connector. At this time, the use of this field is limited to indicating
thread-safety, asynchronous capabilities, and ability to produce native
HDF5 files. Supported flags can be found in {\tt H5VLconnector.h}.

\begin{lstlisting}
/* Capability flags for connector */
#define H5VL_CAP_FLAG_NONE         0    /* No special connector capabilities */
#define H5VL_CAP_FLAG_THREADSAFE   0x01 /* Connector is threadsafe */
#define H5VL_CAP_FLAG_ASYNC        0x02 /* Connector performs operations asynchronously*/
#define H5VL_CAP_FLAG_NATIVE_FILES 0x04 /* Connector produces native file format */
\end{lstlisting}

@@

You'll need to decide how to perform any initialization and shutdown tasks that
are required by your connector. There are initialize and terminate callbacks
in the {\tt H5VL\_class\_t} struct to handle this. They are invoked when
the connector is registered and unregistered, respectively. The inialize
callback can take a VOL initialization property list, so any properties you
need for initialization can be applied to it. The HDF5 library currently makes
no use of the vipl so there are no default vipl properties.

If this is unsuitable, you may have to create custom connector-specific API calls to handle
initialization and termination. It may also be useful to perform operations in
a custom API call used to set the VOL connector in the fapl.

The initialization and terminate callbacks:
\begin{lstlisting}
herr_t (*initialize)(hid_t vipl_id); /**< Connector initialization callback */

herr_t (*terminate)(void);           /**< Connector termination callback */
\end{lstlisting}

@@

Once you have your data structures, you'll need to create your own
implementations of the callback functions and map them via your
{\tt H5VL\_class\_t} struct.

\subsection{Handling Optional Operations}

Handling optional operations has changed significantly in HDF5 1.13.0. In the
past, optional operations were specified using an integer {\tt opt\_type}
parameter. This proved to be a problem with pass-through connectors, though,
as it was possible to have {\tt opt\_type} clash if two connectors used
the same {\tt opt\_type} values.

The new scheme allows a connector to register an optional operation with
the library and receive a dynamically-allocated {\tt opt\_type} value
for the operation.

The following API calls can be used to manage the optional operations:
\begin{lstlisting}
herr_t H5VLregister_opt_operation(H5VL_subclass_t subcls, const char *op_name, int *op_val);
herr_t H5VLfind_opt_operation(H5VL_subclass_t subcls, const char *op_name, int *op_val);
herr_t H5VLunregister_opt_operation(H5VL_subclass_t subcls, const char *op_name);
\end{lstlisting}

The {\tt register} call is used to register an operation for a subclass (file, etc.)
and the {\tt opt\_type} parameter that the library assigned to the operation will be
returned via the {\tt opt\_val} parameter. This value can then be passed to one
of the subclass-specific API calls (listed below). If you need to find an existing
optional call's assigned {\tt opt\_type} value by name, you can use the {\tt find}
call.

One recommended way to handle optional calls is to register all the optional calls
at startup, saving the values in connector state, then use these cached values
in your optional calls. The assigned values should be unregistered using the
{\tt unregister} call when the connector shuts down.

Subclass-specific optional calls:
\begin{lstlisting}
herr_t H5VLattr_optional_op(const char *app_file, const char *app_func, unsigned app_line,
                            hid_t attr_id, H5VL_optional_args_t *args, hid_t dxpl_id, hid_t es_id);
herr_t H5VLdataset_optional_op(const char *app_file, const char *app_func, unsigned app_line,
                               hid_t dset_id, H5VL_optional_args_t *args, hid_t dxpl_id, hid_t es_id);
herr_t H5VLdatatype_optional_op(const char *app_file, const char *app_func, unsigned app_line,
                                hid_t type_id, H5VL_optional_args_t *args, hid_t dxpl_id, hid_t es_id);
herr_t H5VLfile_optional_op(const char *app_file, const char *app_func, unsigned app_line,
                            hid_t file_id, H5VL_optional_args_t *args, hid_t dxpl_id, hid_t es_id);
herr_t H5VLgroup_optional_op(const char *app_file, const char *app_func, unsigned app_line,
                             hid_t group_id, H5VL_optional_args_t *args, hid_t dxpl_id, hid_t es_id);
herr_t H5VLlink_optional_op(const char *app_file, const char *app_func, unsigned app_line,
                            hid_t loc_id, const char *name, hid_t lapl_id, H5VL_optional_args_t *args,
                            hid_t dxpl_id, hid_t es_id);
herr_t H5VLobject_optional_op(const char *app_file, const char *app_func, unsigned app_line,
                              hid_t loc_id, const char *name, hid_t lapl_id,
                              H5VL_optional_args_t *args, hid_t dxpl_id, hid_t es_id);
herr_t H5VLrequest_optional_op(void *req, hid_t connector_id, H5VL_optional_args_t *args);
\end{lstlisting}

\subsection{Testing Your Connector}

@@

\quad \quad \url{https://github.com/HDFGroup/vol-tests}

@@

\subsection{Asynchronous Operations}
